Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-22 paper/paper.tex
start work on examples section

\section{Type-theoretic Characterizations of Language Classes}
Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.

@@

\subsection{TODO More Complicated Grammars}
Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts

>>>>

\section{Type Theoretic Formal Grammar Theory}

Now that we have developed a type theoretic syntax for formal grammars
and parse transformers, we show how many common concepts of formal
language theory and grammars are naturally expressed using type
theoretic notions, and some constructions and proofs can be carried
out in our type theory as well.

Firstly, regular expressions and their extension to context-free
expressions are naturally expressed as a limited syntax of linear
types. That is, regular expressions are precisely the linear types
constructed using only $c, I, \otimes, 0, \oplus$ and the Kleene star,
which can be defined as $A^* = \mu x. I \oplus A \otimes X$. The
context-free expressions are similarly expressible by allowing not
just the Kleene star, but unrestricted usage of $\mu$. Remember,
however, that these are a notion of regular expressions that maintain
a notion of ambiguous grammar. That is, $A \oplus A$ is not considered
to be the same as $A$. While there does not appear to be an analogous
weakening of this definition to naturally describe the
context-sensitive languages, we will see shortly that the indexing of
linear types by non-linear types allows for capturing a notion of
context in grammars.

In addition to expressing algebraic presentations of grammars, the
type-theoretic syntax allows for naturally expressing grammars that
capture the behavior of automata. \max{sync this up with Steven's
  stuff, and add stuff about PDAs and Turing Machines}.

A key construct in formal language theory is the Brzozowski derivative
\cite{brzozowski}. Given a string $w$ the (left) derivative of a
language $L$ with respect to $w$, written $\partial_w L$ is the language
\[ (\partial_wL ) = \{ ww' \pipe w' \in L \}\]
This semantics is given using the linear implication $w \tol
L$. Expressing Brzozowski's calculation of the derivative of a regular
expression relies on additional axioms to the type theory.\max{axioms
  to carry out Brzozowski's derivative calculations plz.}

Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.

@@

\subsection{Indexed Automata}

We can generalize to more powerful notions of automata by utilizing
dependency of linear types on non-linear types. Let $S$ be a finite
non-linear type and define $\textrm{List}\,S = \mu x. 1 + S \times
x$. Then a pushdown automaton with $S$ as the stack alphabet can be
defined as a mutual fixed point.

Want to put PDA, CFGs, CSGs here. Basically, use this section to highlight the usefulness of the intuitionistic part of the contexts
